\documentclass[12pt, leqno]{article} %% use to set typesize
\include{common}

\begin{document}

\pagestyle{fancy}
\lhead{Bindel, Fall 2019}
\rhead{Matrix Computation}
\fancyfoot{}
\begin{center}
  {\large{\bf Final}} \\ (due: 2019-12-18)
\end{center}

You may (and should) use any references you wish: books, notes,
\matlab\ help, literature.  But please do not consult with anyone
outside the course staff.  Templates for the required codes are
provided in the class Github repository.

% Least squares with regularization, constraints
% Eigenvalue problems
% Symmetric eigenproblems
% Iterative methods

\paragraph*{1: Stationary steps}
Consider the iteration
\[
  x^{(k+1)} = x^{(k)} + M^{-1} A^T (b-Ax^{(k)}) 
\]
where $M$ is positive definite and $A \in \bbR^{m \times n}$ with $m
\geq n$ is full rank.
\begin{itemize}
\item[2 pts] What is the fixed point of the iteration?
\item[2 pts] What is the matrix for the error iteration?
\end{itemize}

\paragraph*{2: Eigenbounds}
Let $A \in \bbR^{n \times n}$ be symmetric and
$E \in \bbR^{n \times n}$ be arbitrary, but with $\|E\|_2 \leq
\epsilon$.
\begin{itemize}
\item[2 pts] Argue that $(A-sI) + E$ must be nonsingular if
  $|s-\lambda| > \epsilon$ for every eigenvalue $\lambda$ of $A$.
\item[2 pts] Argue that the eigenvalues of $A+E$ must lie in a union
  of disks of radius $\epsilon$ about the eigenvalues of $A$.
\item[2 pts] Argue that if all the eigenvalues of $A$ are at least $2
  \epsilon$ apart, then all the eigenvalues of $A+E$ must be real.
\end{itemize}

\paragraph*{3: Eigenvector estimation}
Let the columns of $V \in \bbR^{n \times k}$ be an orthonormal basis
for a $k$-dimensional subspace $\mathcal{V} \subset \bbR^n$.
Suppose $A \in \bbR^{n \times n}$ has an approximate eigenvalue $\mu
\in \bbR$ (given)
\begin{itemize}
\item[2 pts] Write an efficient code to find a unit-length vector
  $v \in \mathcal{V}$ such that $\|(A-\mu I)v\|$ is minimal.
\item[2 pts] Give a backward error bound for the $v$ computed in
  the previous part (i.e.~give a bound on the norm of an $E$ such that
  $(A+E) v = \mu v$).
\end{itemize}

\paragraph*{4: Concatenated constraints}
Consider the problem
\[
  \mbox{minimize } x^T A x \mbox{ s.t. } x^T M x = 1 \mbox{ and } Cx = b
\]
where $A, M \in \bbR^{n \times n}$ are symmetric, $M$ is positive definite, and $C \in
\bbR^{m \times n}$ with $m < n$.
\begin{itemize}
\item[2 pts] Find $\bar{x}$ that minimizes $x^T M x$ subject to $Cx = b$.
\item[2 pts] Show how to compute an $M$-orthonormal basis $V$ for the 
  null space of $C$ (this can be a MATLAB or Julia fragment).
\item[2 pts] Argue that $x = \bar{x} + Vy$ for some $y$,
  and show that $y$ can be computed via an eigenvalue problem
  (yoou may refer to HW solutions from class).
\end{itemize}


\end{document}
