\documentclass[12pt, leqno]{article} %% use to set typesize
\include{common}

\begin{document}

\pagestyle{fancy}
\lhead{Bindel, Fall 2019}
\rhead{Matrix Computation}
\fancyfoot{}
\begin{center}
  {\large{\bf Midterm}} \\ (due: 2019-10-21)
\end{center}

You may (and should) use any references you wish: books, notes,
\matlab\ help, literature.  But please do not consult with anyone
outside the course staff.  Templates for the required codes are
provided in the class Github repository.

\paragraph*{1: Identity plus}
Consider the matrix $A = I + ZZ^T$ where $Z \in \bbR^{n \times k}$,
$k \ll n$.
\begin{itemize}
\item[2 pts] Argue that $A$ is symmetric and positive definite.
\item[2 pts] Give an $O(nk)$ time algorithm for computing $y = Ax$.
\item[2 pts] Given an economy SVD $Z = U \Sigma V^T$, show how to compute
  $\kappa_2(A)$ cheaply.  How cheaply can it be done?
\item[2 pts] Given an economy QR decomposition $Z = QR$, show how to solve
  $Ax = b$ cheaply.  How cheaply can it be done?
\end{itemize}
Please provide {\em both} a written description of your approach and
code that satisfies the provided interfaces.

\paragraph*{2: Daring derivatives}
Give codes to compute directional derivatives for each of the
following
\begin{itemize}
\item[2 pts] Differentiate $\|x\|_M^2$ with respect to changes in $x$ and
  changes in $M$ (for $M$ spd).
\item[2 pts] Differentiate the solution to $(I+ZZ^T) x = b$ with respect to
  changes in $Z$ and $b$, given a QR factorization of $Z$.  Your code
  should require $O(nk + k^3)$ time.
\item[2 pts] Differentiate the Cholesky factorization $R^T R = M$
  ({\em Hint}: differentiate the basic relation, then pre- and
  post-multiply by $R^{-T}$ and $R^{-1}$, respectively).
\end{itemize}
Please provide {\em both} a written description of your approach and
code that satisfies the provided interfaces.

\paragraph*{3: Tridiagonal trouble}
Let $T$ be a positive definite symmetric tridiagonal matrix
\[
T = \begin{bmatrix}
  \alpha_1 & \beta_1 \\
  \beta_1 & \alpha_2 & \beta_2 \\
  & \beta_2 & \alpha_2 & \beta_3 \\
  & & \ddots & \ddots & \ddots \\
  & & & \beta_{n-2} & \alpha_{n-1} &\beta_{n-1} \\
  & & & & \beta_{n-1} \alpha_n
  \end{bmatrix}
\]
Answer the following question
\begin{itemize}
\item[2 pts] Consider any block 2-by-2 decomposition of $S = T^{-1}$ with
  square diagonal blocks, and show that $S_{12}$ and $S_{21}$ are rank
  one.
\item[2 pts] Write a code to compute $\|T^{-1}\|_1$ efficiently and exactly
  (i.e.~you should not use a technique like Hager's condition estimator).
  Please provide {\em both} a written description of your approach and
  code that satisfies the provided interfaces.
\end{itemize}

\paragraph*{4: Conditioning}
[2 pts] For any operator norm, show that if $\|A^{-1} E\| < 1$ then
\[
  \kappa(A+E)
  \leq \kappa(A) \left( \frac{1+\|A^{-1} E\|}{1-\|A^{-1}E\|} \right)
  \leq \frac{\kappa(A)}{(1-\|A^{-1}E\|)^2}
\]
where $\kappa$ denotes the condition number with respect to linear solves.

\end{document}
