\documentclass[12pt, leqno]{article} %% use to set typesize
\include{common}

\begin{document}

\hdr{2019-09-09}

\section{Matrix calculus}

Numerical linear algebra is not just about algebra, but also about
{\em analysis}, the branch of mathematics that deals with real
functions and operations such as differentiation and integration
(and their generalizations).  This is particularly relevant when
we deal with error analysis, as we will soon.

\subsection{Warm up: derivative of a dot product}

Consider the real-valued expression $y^T x$ as a function of
the vector variables $x, y \in \bbR^n$.  How would we compute
the gradient of $y^T x$ with respect to these variables?
The usual method taught in a first calculus class would be
to write the expression in terms of each of the components of $x$
and $y$, and then compute partial derivatives, i.e.
\begin{align*}
  y^T x &= \sum_{i=1}^n x_i y_i \\
  \frac{\partial (y^T x)}{\partial x_j} &= y_j \\
  \frac{\partial (y^T x)}{\partial y_j} &= x_j.
\end{align*}
This notation is fine for dealing with a dot product, in which
we are summing over only one variable; but when we deal with more
complicated matrix expressions, it quickly becomes painful to deal
with coordinates.  A neat trick of notation is to work not with
derivatives along the coordinate directions, but with derivatives
in an {\em arbitrary} direction
$(\delta x, \delta y) \in \bbR^n \times \bbR^n$:
\[
  \left. \frac{d}{ds} \right|_{s=0} (y+s \delta y)^T (x + s \delta x)
  = \delta y^T x + y^T \delta x.
\]
We denote the directional derivative by $\delta(y^T x)$, giving the
tidy expression
\[
  \delta(y^T x) = \delta y^T x + y^T \delta x.
\]
This is {\em variational notation} for reasoning about directional
(Gateaux) derivatives.  It is often used in
mechanics and in PDE theory and functional analysis
(where the vector spaces involved are infinite-dimensional),
and I have always felt it deserves to be used more widely.

\subsection{Some calculus facts}

We will make frequent use of the humble product rule in this class:
\[
  \delta (AB) = \delta A \, B + A \, \delta B.
\]
As is always the case, the order of the terms in the products is
important.  To differentiate a product of three terms (for example),
we would have
\[
  \delta (ABC) = (\delta A) BC + A (\delta B) C + AB(\delta C).
\]

The product rule and implicit differentiation gives us
\[
  0 = \delta (A^{-1} A) = \delta (A^{-1}) A + A^{-1} \delta A.
\]
Rearranging slightly, we have
\[
  \delta (A^{-1}) = -A^{-1} (\delta A) A^{-1},
\]
which is again a matrix version of the familiar rule from Calculus I,
differing only in that we have to be careful about the order of
products.  This rule also nicely illustrates the advantage of variational
notation; if you are unconvinced, I invite you to write out the
elements of the derivative of a matrix inverse using conventional
coordinate notation!

The vector 2-norm and the Frobenius norm for matrices are convenient
because the (squared) norm is a differentiable function of the entries.
For the vector 2-norm, we have
\[
  \delta(\|x\|^2) = \delta(x^* x) = (\delta x)^* x + x^* (\delta x);
\]
observing that $y^* x = (x^* y)^*$ and $z + \bar{z} = 2 \Re(z)$,
we have
\[
  \delta(\|x\|^2) = 2 \Re(\delta x^* x).
\]
Similarly, the Frobenius norm is associated with a dot product
(the unsurprisingly-named Frobenius inner product) on all the elements
of the matrix, which we can write in matrix form as
\[
  \ip{A}{B}_F = \tr(B^* A),
\]
and we therefore have
\[
  \delta (\|A\|_F^2) = \delta \tr(A^* A) = 2 \Re \tr(\delta A^* A).
\]

\subsection{The 2-norm revisited}

In the previous lecture, we discussed the matrix 2-norm in terms
of the singular value decomposition.  What if we did not know about
the SVD?  By the definition, we would like to maximize
$\phi(v)^2 = \|Av\|^2$ subject to $\|v\|^2 = 1$.  Flexing our
new variational notation, let's work through the first-order condition
for a maximum.  To enforce the condition, we form an augmented
Lagrangian
\[
  L(v,\mu) = \|Av\|^2 - \mu (\|v\|^2 - 1)
\]
and differentiating gives us
\[
  \delta L = 2 \Re (\delta v^* (A^* A v - \mu v)) - \delta \mu (\|v\|^2 - 1).
\]
The first-order condition for a maximum or minimum is $\delta L = 0$
for all possible $\delta v$ and $\delta \mu$; this gives
\[
  A^* A v = \mu v, \quad \|v\|^2 = 1,
\]
which is an eigenvalue problem involving the Gram matrix $A^* A$.
We will see this eigenvalue problem again --- and the more general idea
of the connection between eigenvalue problems and
optimizing quadratic forms --- later in the course.

\subsection{Norms and Neumann series}

We will do a great deal of operator norm manipulation this semester,
almost all of which boils down to repeated use of the triangle
inequality and the submultiplicative property.  For now, we illustrate
the point by a simple, useful example: the matrix version of
the geometric series.

Suppose $F$ is a square matrix such that $\|F\| < 1$ in
some operator norm, and consider the power series
\[
  \sum_{j=0}^n F^j.
\]
Note that $\|F^j\| \leq \|F\|^j$ via the submultiplicative property
of induced operator norms.
By the triangle inequality, the partial sums satisfy
\[
  (I-F) \sum_{j=0}^n F^j = I - F^{n+1}.
\]
Hence, we have that
\[
  \|(I-F) \sum_{j=0}^n F^j - I\| \leq \|F\|^{n+1} \rightarrow 0
  \mbox{ as } n \rightarrow \infty,
\]
i.e. $I-F$ is invertible and the inverse is given by the convergent
power series (the geometric series or {\em Neumann series})
\[
  (I-F)^{-1} = \sum_{j=0}^\infty F^j.
\]
By applying submultiplicativity and triangle inequality to the partial
sums, we also find that
\[
  \|(I-F)^{-1}\| \leq \sum_{j=0}^\infty \|F\|^j = \frac{1}{1-\|F\|}.
\]

Note as a consequence of the above that if $\|A^{-1} E\| < 1$ then
\[
  \|(A+E)^{-1}\|
  = \|(I+A^{-1} E)^{-1} A^{-1}\|
  \leq \frac{\|A^{-1}\|}{1-\|A^{-1} E\|}.
\]
That is, the Neumann series gives us a sense of how a small
perturbation to $A$ can change the norm of $A^{-1}$.

\end{document}
