\documentclass[12pt, leqno]{article} %% use to set typesize
\include{common}

\begin{document}

\hdr{2019-09-16}

\section{Finding and fixing floating point problems}

Floating point arithmetic is not the same as real arithmetic.  Even
simple properties like associativity or distributivity of addition and
multiplication only hold approximately.  Thus, some computations that
look fine in exact arithmetic can produce bad answers in floating
point.  What follows is a (very incomplete) list of some of the ways
in which programmers can go awry with careless floating point
programming.

\subsection{Cancellation}

If $\hat{x} = x(1+\delta_1)$ and $\hat{y} = y(1+\delta_2)$ are
floating point approximations to $x$ and $y$ that are very close, then
$\fl(\hat{x}-\hat{y})$ may be a poor approximation to $x-y$ due to
{\em cancellation}.  In some ways, the subtraction is blameless in
this tail: if $x$ and $y$ are close, then $\fl(\hat{x}-\hat{y}) =
\hat{x}-\hat{y}$, and the subtraction causes no additional rounding
error.  Rather, the problem is with the approximation error already
present in $\hat{x}$ and $\hat{y}$.

The standard example of loss of accuracy revealed through cancellation
is in the computation of the smaller root of a quadratic using the
quadratic formula, e.g.
\[
  x = 1-\sqrt{1-z}
\]
for $z$ small.  Fortunately, some algebraic manipulation gives an
equivalent formula that does not suffer cancellation:
\[
  x =
    \left( 1-\sqrt{1-z} \right)
    \left(\frac{1+\sqrt{1-z}}{1+\sqrt{1-z}}\right)
  =
    \frac{z}{1+\sqrt{1-z}}.
\]

\subsection{Sensitive subproblems}

We often solve problems by breaking them into simpler subproblems.
Unfortunately, it is easy to produce badly-conditioned subproblems
as steps to solving a well-conditioned problem.  As a simple (if
contrived) example, try running the following MATLAB code:

\lstinputlisting{code/sillysqrt.m}

In exact arithmetic, this should produce 2, but what does it produce
in floating point?  In fact, the first loop produces a correctly
rounded result, but the second loop represents the function
$x^{2^{60}}$, which has a condition number far greater than $10^{16}$
--- and so all accuracy is lost.

\subsection{Unstable recurrences}

One of my favorite examples of this problem is the recurrence relation
for computing the integrals
\[
  E_n = \int_{0}^1 x^n e^{x-1} \, dx.
\]
Integration by parts yields the recurrence
\begin{align*}
  E_0 &= 1-1/e \\
  E_n &= 1-nE_{n-1}, \quad n \geq 1.
\end{align*}
This looks benign enough at first glance: no single step of
this recurrence causes the error to explode.  But
each step amplifies the error somewhat, resulting in an exponential
growth in error\footnote{%
Part of the reason that I like this example is that one can
run the recurrence {\em backward} to get very good results,
based on the estimate $E_n \approx 1/(n+1)$ for $n$ large.
}.

\subsection{Undetected underflow}

In Bayesian statistics, one sometimes computes ratios of long
products.  These products may underflow individually, even when the
final ratio is not far from one.  In the best case, the products will
grow so tiny that they underflow to zero, and the user may notice an
infinity or NaN in the final result.  In the worst case, the
underflowed results will produce nonzero subnormal numbers with
unexpectedly poor relative accuracy, and the final result will be
wildly inaccurate with no warning except for the (often ignored)
underflow flag.

\subsection{Bad branches}

A NaN result is often a blessing in disguise: if you see an
unexpected NaN, at least you {\em know} something has gone wrong!
But all comparisons involving NaN are false,  and so
when a floating point result is used to compute a branch condition
and an unexpected NaN appears, the result can wreak havoc.
As an example, try out the following code in MATLAB.

\lstinputlisting{code/testnan.m}

\section{Sums and dots}

We already described a couple of floating point examples that
involve evaluation of a fixed formula (e.g.~computation of
the roots of a quadratic).  We now turn to the analysis of
some of the building blocks for linear algebraic computations:
sums and dot products.

\subsection{Sums two ways}

As an example of first-order error analysis, consider the following
code to compute a sum of the entries of a vector $v$:
\begin{lstlisting}
  s = 0;
  for k = 1:n
    s = s + v(k);
  end
\end{lstlisting}
Let $\hat{s}_k$ denote the computed sum at step $k$ of the loop;
then we have
\begin{align*}
  \hat{s}_1 &= v_1 \\
  \hat{s}_k &= (\hat{s}_{k-1} + v_k)(1 + \delta_k), \quad k > 1.
\end{align*}
Running this forward gives
\begin{align*}
  \hat{s}_2 &= (v_1 + v_2)(1+\delta_2) \\
  \hat{s}_3 &= ((v_1 + v_2)(1+\delta_2) + v_3)(1+\delta_2)
\end{align*}
and so on.  Using first-order analysis, we have
\[
  \hat{s}_k \approx (v_1 + v_2)\left(1 + \sum_{j=2}^k \delta_j \right)
              + \sum_{l=3}^k v_l \left( 1 + \sum_{j=l}^k \delta_j \right),
\]
and the difference between $\hat{s}_k$ and the exact partial sum
is then
\[
  \hat{s}_k-s_k \approx \sum_{j=2}^k s_j \delta_j.
\]
Using $\|v\|_1$ as a uniform bound on all the partial sums, we have
\[
  |\hat{s}_n-s_n| \lesssim (n-1) \macheps \|v\|_2.
\]

An alternate analysis, which is a useful prelude to analyses to come
involves writing an error recurrence.
Taking the difference between $\hat{s}_k$ and the true partial sums $s_k$,
we have
\begin{align*}
  e_1 &= 0 \\
  e_{k} &= \hat{s}_k-s_k \\
        &= (\hat{s}_{k-1} + v_k)(1+\delta_k) - (s_{k-1} + v_k) \\
        &= e_{k-1} + (\hat{s}_{k-1} + v_k) \delta_k,
\end{align*}
and $\hat{s}_{k-1} + v_k = s_k + O(\macheps)$, so that
\[
  |e_{k}| \leq |e_{k-1}| + |s_k| \macheps + O(\macheps^2).
\]
Therefore,
\[
  |e_{n}| \lesssim (n-1) \macheps \|v\|_1,
\]
which is the same bound we had before.

\subsection{Backward error analysis for sums}

In the previous subsection, we showed an error analysis for partial sums
leading to the expression:
\[
  \hat{s}_n \approx (v_1 + v_2) \left(1 + \sum_{j=2}^n \delta_j \right)
              + \sum_{l=3}^n v_l \left( 1 + \sum_{j=l}^n \delta_j \right).
\]
We then proceded to aggregate all the rounding error terms in order to
estimate the error overall.  As an alternative to aggregating the roundoff,
we can also treat the rounding errors as perturbations to the input
variables (the entries of $v$); that is, we write the computed sum as
\[
  \hat{s}_n = \sum_{j=1}^n \hat{v}_j
\]
where
\[
  \hat{v}_j = v_j (1+\eta_j), \quad
  \mbox{ where }|\eta_j| \lesssim (n+1-j) \macheps.
\]
This gives us a {\em backward error} formulation of the rounding:
we have re-cast the role of rounding error in terms of a
perturbation to the input vector $v$.  In terms of the 1-norm,
we have the relative error bound
\[
  \|\hat{v}-v\|_1 \lesssim n \macheps \|v\|_1;
\]
or we can replace $n$ with $n-1$ by being a little more careful.
Either way, what we have shown is that the summation algorithm
is {\em backward stable}, i.e.~we can ascribe the roundoff to
a (normwise) small relative error with a bound of $C \macheps$
where the constant $C$ depends on the size $n$ like some low-degree
polynomial.

Once we have a bound on the backward error, we can bound the forward
error via a condition number.  That is, suppose we write the true
and perturbed sums as
\begin{align*}
  s &= \sum_{j=1}^n v_j &
  \hat{s} &= \sum_{j=1}^n \hat{v}_j.
\end{align*}
We want to know the relative error in $\hat{s}$ via a normwise
relative error bound in $\hat{v}$, which we can write as
\[
  \frac{|\hat{s}-s|}{|s|} =
  \frac{|\sum_{j=1}^n (\hat{v}_j-v_j)|}{|s|} \leq
  \frac{\|\hat{v}-v\|_1}{|s|} =
  \frac{\|v\|_1}{|s|} \frac{\|\hat{v}-v\|_1}{\|v\|_1}.
\]
That is, $\|v\|_1/|s|$ is the condition number for the summation
problem, and our backward stability analysis implies
\[
  \frac{|\hat{s}-s|}{|s|} \leq
  \frac{\|v\|_1}{|s|} n \macheps.
\]
This is the general pattern we will see again in the future:
our analysis consists of a backward error computation that depends
purely on the algorithm, together with a condition number that depends
purely on the problem.  Together, these give us forward error bounds.

\subsection{Running error bounds for sums}

In all the analysis of summation we have done so far, we ultimately
simplified our formulas by bounding some quantity in terms of $\|v\|_1$.
This is nice for algebra, but we lose some precision in the process.
An alternative is to compute a {\em running error bound}, i.e.~augment
the original calculation with something that keeps track of the error
estimates.  We have already seen that the error in the computations
looks like
\[
  \hat{s}_n-s_n = \sum_{j=2}^n s_j \delta_j + O(\macheps^2),
\]
and since $s_j$ and $\hat{s}_j$ differ only by $O(\macheps)$ terms,
\[
  |\hat{s}_n-s_n| \lesssim \sum_{j=2}^n |\hat{s}_j| \macheps + O(\macheps^2),
\]
We are not worried about doing a rounding error analysis of our rounding
error analysis --- in general, we care more about order of magnitude for
rounding error anyhow --- so the following routine does an adequate job
of computing an (approximate) upper bound on the error in the summation:
\begin{lstlisting}
  s = 0;
  e = 0;
  for k = 1:n
    s = s + v(k);
    e = e + abs(s) * eps;
  end
\end{lstlisting}

\subsection{Compensated summation}

We conclude our discussion of rounding analysis for summation
with a comment on the {\em compensated summation} algorithm of
Kahan, which is not amenable to straightforward $1+\delta$ analysis.
The algorithm maintains the partial sums not as a single variable
{\tt s}, but as an unevaluated sum of two
variables {\tt s} and {\tt c}:
\begin{lstlisting}
  s = 0;
  c = 0;
  for k = 1:n
    y = v(i) - c;
    t = s + y;
    c = (t - s) - y;  % Key step
    s = t;
  end
\end{lstlisting}
Where the error bound for ordinary summation is
$(n-1) \macheps \|v\|_1 + O(\macheps^2)$,
the error bound for compensated summation is
$2 \macheps \|v\|_1 + O(\macheps^2)$.  Moreover,
compensated summation is exact for adding up to $2^k$
terms that are within about $2^{p-k}$ of each other
in magnitude.

Nor is Kahan's algorithm the end of the story!
Higham's {\em Accuracy and Stability of Numerical Methods}
devotes an entire chapter to summation methods, and there
continue to be papers written on the topic.  For our purposes,
though, we will wrap up here with two observations:
\begin{itemize}
\item Our initial analysis in the $1+\delta$ model illustrates the
  general shape these types of analyses take and how we can re-cast
  the effect of rounding errors as a ``backward error'' that perturbs
  the inputs to an exact problem.
\item The existence of algorithms like Kahan's compensated summation
  method should indicate that the backward-error-and-conditioning
  approach to rounding analysis is hardly the end of the story.
  One could argue it is hardly the beginning!  But it is the approach
  we will be using for most of the class.
\end{itemize}

\subsection{Dot products}

We conclude with one more example error analysis, this time involving
a real dot product computed by a loop of the form
\begin{lstlisting}
  dot = 0;
  for k = 1:n
    dot = dot + x(k)*y(k);
  end
\end{lstlisting}
Unlike the simple summation we analyzed above, the dot product involves
two different sources of rounding errors: one from the summation,
and one from the product.  As in the case of simple summations, it is
convenient to re-cast this error in terms of perturbations to the input.
We could do this all in one go, but since we have already spent so much
time on summation, let us instead do it in two steps.  Let $v_k = x_k y_k$;
in floating point, we get $\hat{v}_k = v_k (1+\eta_k)$
where $|\eta_k| < \macheps$.  Further,
we have already done a backward error analysis of summation to show
that the additional error in summation can be cast onto the summands,
i.e.~the floating point result is $\sum_k \tilde{v}_k$ where
\begin{align*}
  \tilde{v}_k
  &= \hat{v}_k (1 + \sum_{j=\min(2,n)}^n \delta_j) (1 + \eta_k) + O(\macheps^2) \\
  &= v_k (1+\gamma_k) + O(\macheps^2)
\end{align*}
where
\[
  |\gamma_k| = |\eta_k + \sum_{j=\min(2,n)}^n \delta_j| \leq n \macheps.
\]
Rewriting $v_k(1+\gamma_k)$ as $\hat{x}_k y_k$ where
$\hat{x}_k = x_k (1+\gamma_k)$, we have that the computed inner
product $y^T x$ is equivalent to the exact inner product of $y^T \hat{x}$
where $\hat{x}$ is an elementwise relatively accurate
(to within $n \macheps + O(\macheps^2)$) approximation to $x$.

A similar backward error analysis shows that computed matrix-matrix
products $AB$ in general can be interpreted as $\hat{A} B$ where
\[
  |\hat{A}-A| < p \macheps |A| + O(\macheps^2)
\]
and $p$ is the inner dimension of the product.  Exactly what $\hat{A}$
is depends not only on the data, but also the loop order used in the
multiply --- since, as we recall, the order of accumulation may vary
from machine to machine depending on what blocking is best suited to
the cache!  But the bound on the backward error holds for all the common
re-ordering\footnote{%
For those of you who know about Strassen's algorithm --- it's not
backward stable, alas.
}
And this backward error characterization, together with the
type of sensitivity analysis for matrix multiplication that we have
already discussed, gives us a uniform framework for obtaining
forward error bounds for matrix-matrix muliplication; and the same
type of analysis will continue to dominate our discussion of rounding
errors as we move on to more complicated matrix computations.

\section{Problems to ponder}

\begin{enumerate}
\item
  How do we accurately evaluate $\sqrt{1+x}-\sqrt{1-x}$ when $x \ll 1$?
\item
  How do we accurately evaluate $\ln \sqrt{x+1}-\ln\sqrt{x}$ when $x \gg 1$?
\item
  How do we accurately evaluate $(1-\cos(x))/\sin(x)$ when $x \ll 1$?
\item
  How would we compute $\cos(x)-1$ accurately when $x \ll 1$?
\item
  The {\em Lamb-Oseen vortex} is a solution to the 2D Navier-Stokes
  equation that plays a key role in some methods for computational
  fluid dynamics.  It has the form
  \[
    v_{\theta}(r,t) = \frac{\Gamma}{2 \pi r}
      \left( 1 - \exp\left( \frac{-r^2}{4 \nu t} \right) \right)
  \]
  How would one evaluate $v(r,t)$ to high relative accuracy for all
  values of $r$ and $t$ (barring overflow or underflow)?
\item
  For $x > 1$, the equation $x = \cosh(y)$ can be solved as
  \[
    y = -\ln\left(x - \sqrt{x^2-1}\right).
  \]
  What happens when $x = 10^8$?  Can we fix it?
\item
  The difference equation
  \[
    x_{k+1} = 2.25x_k - 0.5 x_{k-1}
  \]
  with starting values
  \[
    x_1 = \frac{1}{3}, \qquad x_2 = \frac{1}{12}
  \]
  has solution
  \[
    x_k = \frac{4^{1-k}}{3}.
  \]
  Is this what you actually see if you compute?  What goes wrong?
\item
  Considering the following two MATLAB fragments:
\begin{lstlisting}
    % Version 1
    f = (exp(x)-1)/x;

    % Version 2
    y = exp(x);
    f = (1-y)/log(y);
\end{lstlisting}
  In exact arithmetic, the two fragments are equivalent.  In floating
  point, the first formulation is inaccurate for $x \ll 1$, while the
  second formulation remains accurate.  Why?
\item
  Running the recurrence $E_n = 1-nE_{n-1}$ {\em forward} is an
  unstable way to compute $\int_0^1 x^n e^{x-1} \, dx$.  However, we
  can get good results by running the recurrence {\em backward} from
  the estimate $E_n \approx 1/(N+1)$ starting at large enough $N$.
  Explain why.  How large must $N$ be to compute $E_{20}$ to near
  machine precision?
\item
  How might you accurately compute this function for $|x| < 1$?
  \[
    f(x) = \sum_{j=0}^\infty \left( \cos(x^j)-1 \right)
  \]
\end{enumerate}

\end{document}
